---
title: "_HDF5Array_ performance"
author:
- name: Hervé Pagès
  affiliation: Fred Hutch Cancer Center, Seattle, WA
date: "Compiled `r doc_date()`; Modified 15 January 2024"
package: HDF5Array
vignette: |
  %\VignetteIndexEntry{HDF5Array performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document
---


# Introduction

The aim of this short document is to measure the performance of
the `r Biocpkg("HDF5Array")` package for normalization and PCA,
two operations commomly involved in the context of single cell analysis.

The goal is to facilitate comparison with other tools like Seurat
and Scanpy.


# Install and load required packages

Let's install and load `r Biocpkg("HDF5Array")` as well as the other
packages used in this vignette:
```{r, eval=FALSE}
if (!require("BiocManager", quietly=TRUE))
    install.packages("BiocManager")

pkgs <- c("HDF5Array", "ExperimentHub", "DelayedMatrixStats", "RSpectra")
BiocManager::install(pkgs)
```

Load the package:
```{r, message=FALSE}
library(HDF5Array)
library(ExperimentHub)
library(DelayedMatrixStats)
library(RSpectra)
```


# The test datasets

## _Sparse_ vs _dense_ representation

The datasets that we will use in this document are subsets of the
_1.3 Million Brain Cell Dataset_ from 10x Genomics.

The _1.3 Million Brain Cell Dataset_ is a 27,998 x 1,306,127 matrix
of counts with one gene per row and one cell per column. It's available
via the `r Biocpkg("ExperimentHub")` package in two forms, one that uses
a sparse representation and one that uses a dense representation:
```{r}
hub <- ExperimentHub()
hub["EH1039"]$description  # sparse representation
hub["EH1040"]$description  # dense representation
```

The two datasets are big HDF5 files stored on a remote location.
Let's download them to the local `r Biocpkg("ExperimentHub")` cache
if they are not there yet:
```{r, message=FALSE}
## Note that this will be quick if the HDF5 files are already in the
## local ExperimentHub cache. Otherwise, it will take a while
full_sparse_h5 <- hub[["EH1039"]]
full_dense_h5  <- hub[["EH1040"]]
```

## TENxMatrix vs HDF5Matrix object

We use the `TENxMatrix()` and `HDF5Array()` constructors to bring the
sparse and dense datasets in R as DelayedArray derivatives. Note that
this does _not_ load the matrix data in memory.

Bring the sparse dataset in R:
```{r}
## Use 'h5ls(full_sparse_h5)' to find out the group.
full_sparse <- TENxMatrix(full_sparse_h5, group="mm10")
class(full_sparse)
dim(full_sparse)
```

Bring the dense dataset in R:
```{r}
## Use 'h5ls(full_dense_h5)' to find out the name of the dataset.
full_dense <- HDF5Array(full_dense_h5, name="counts")
class(full_dense)
dim(full_dense)
```

Note that the dense HDF5 file does not contain the dimnames of the
matrix so we manually add them:
```{r}
dimnames(full_dense) <- dimnames(full_sparse)
```

## Create the test datasets

For our benchmarks below we'll use subsets of the _1.3 Million Brain
Cell Dataset_ of increasing sizes: subsets with 12,500 cells, 25,000 cells,
50,000 cells, 100,000 cells, and 200,000 cells:
```{r}
sparse1 <- full_sparse[ , 1:12500]
dense1  <- full_dense[ , 1:12500]

sparse2 <- full_sparse[ , 1:25000]
dense2  <- full_dense[ , 1:25000]

sparse3 <- full_sparse[ , 1:50000]
dense3  <- full_dense[ , 1:50000]

sparse4 <- full_sparse[ , 1:100000]
dense4  <- full_dense[ , 1:100000]

sparse5 <- full_sparse[ , 1:200000]
dense5  <- full_dense[ , 1:200000]
```

# Block-processed normalization and PCA

## Code used for normalization and PCA

We'll use the following code for normalization:
```{r}
## Keep 1000 most variable genes by default.
simple_normalize <- function(mat, num_variable_genes=1000)
{
    stopifnot(length(dim(mat)) == 2, !is.null(rownames(mat)))
    mat <- mat[rowSums(mat) > 0, ]
    mat <- t(t(mat) * 10000 / colSums(mat))
    row_vars <- rowVars(mat)
    rv_order <- order(row_vars, decreasing=TRUE)
    variable_idx <- head(rv_order, n=num_variable_genes)
    mat <- log1p(mat[variable_idx, ])
    mat / rowSds(mat)
}
```

and the following code for PCA:
```{r}
simple_PCA <- function(mat, k=25)
{
    stopifnot(length(dim(mat)) == 2)
    row_means <- rowMeans(mat)
    Ax <- function(x, args)
        (as.numeric(mat %*% x) - row_means * sum(x))
    Atx <- function(x, args)
        (as.numeric(x %*% mat) - as.vector(row_means %*% x))
    RSpectra::svds(Ax, Atrans=Atx, k=k, dim=dim(mat))
}
```

## Block processing and block size

Note that the implementations of `simple_normalize()` and `simple_PCA()`
are expected to work on any matrix-like object regardless of its exact
type/representation e.g. it can be an ordinary matrix, a SparseMatrix
object from the `r Biocpkg("SparseArray")` package, a dgCMatrix object
from the `r Biocpkg("Matrix")` package, a DelayedMatrix derivative
(TENxMatrix, HDF5Matrix, TileDBMatrix), etc...

However, when the input is a DelayedMatrix object or derivative,
it's important to be aware that:

- Summarization methods like `sum()`, `colSums()`, `rowVars()`, or `rowSds()`,
  and matrix multiplication (`%*%`), are _block-processed_ operations.

- The block size is 100 Mb by default. Increasing or decreasing the block
  size will increase or decrease the memory usage of _block-processed_
  operations.

- The block size can be controlled with `DelayedArray::getAutoBlockSize()`
  and `DelayedArray::setAutoBlockSize()`.

For out benchmarks below, we'll use the following block sizes:
- normalization of the sparse datasets: 250 Mb
- normalization of the dense datasets: 100 Mb
- PCA on the normalized sparse datasets: 100 Mb
- PCA on the normalized dense datasets: 100 Mb

## Monitoring memory usage

While manually running our benchmarks below on a Linux system, we also
monitor memory usage at the command line in a terminal with:

    (while true; do ps u -p <PID>; sleep 1; done) >ps.log 2>&1 &

where `<PID>` was the process id of our R session. This allowed us to measure
the maximum amount of memory used by the calls to `simple_normalize()`
or `simple_PCA()`.


# Normalization benchmarks

In this section we run `simple_normalize()` on the two smaller test
datasets only (27,998 x 12,500 and 27,998 x 25,000, sparse and dense),
and we report timings and memory usage.

See **Timings observed on various systems** section at the end of this
document for `simple_normalize()` and `simple_pca()` times observed
on all our test datasets on various systems.

## Normalizing the sparse datasets

```{r}
DelayedArray::setAutoBlockSize(2.5e8)  # blocks of 250 Mb
```

### 27,998 x 12,500 sparse dataset

```{r normalize_sparse1}
dim(sparse1)
system.time(sparse1n <- simple_normalize(sparse1))
gc()
```

Saving the normalized dataset to a temporary file for PCA later:
```{r, results="hide"}
dim(sparse1n)
sparse1n_path <- tempfile()
writeTENxMatrix(sparse1n, sparse1n_path, group="matrix", level=0)
```

### 27,998 x 25,000 sparse dataset

```{r normalize_sparse2}
dim(sparse2)
system.time(sparse2n <- simple_normalize(sparse2))
gc()
```

Saving the normalized dataset to a temporary file for PCA later:
```{r, results="hide"}
dim(sparse2n)
sparse2n_path <- tempfile()
writeTENxMatrix(sparse2n, sparse2n_path, group="matrix", level=0)
```

### About memory usage

With this block size (250 Mb), memory usage (as reported by Unix
command `ps u -p <PID>`, see **Monitoring memory usage** above in
this document) remained < 3.7 Gb at all time.

## Normalizing the dense datasets

```{r}
DelayedArray::setAutoBlockSize(1e8)  # blocks of 100 Mb
```

### 27,998 x 12,500 dense dataset

```{r normalize_dense1}
dim(dense1)
system.time(dense1n <- simple_normalize(dense1))
gc()
```

Saving the normalized dataset to a temporary file for PCA:
```{r, results="hide"}
dim(dense1n)
dense1n_path <- tempfile()
writeHDF5Array(dense1n, dense1n_path, name="normalized_counts", level=0)
```

### 27,998 x 25,000 dense dataset

```{r normalize_dense2}
dim(dense2)
system.time(dense2n <- simple_normalize(dense2))
gc()
```

Saving the normalized dataset to a temporary file for PCA:
```{r, results="hide"}
dim(dense2n)
dense2n_path <- tempfile()
writeHDF5Array(dense2n, dense2n_path, name="normalized_counts", level=0)
```

### About memory usage

With this block size (100 Mb), memory usage (as reported by Unix
command `ps u -p <PID>`, see **Monitoring memory usage** above in
this document) remained < 2.8 Gb at all time.


# PCA benchmarks

In this section we run `simple_pca()` on the two normalized datasets
obtained in the previous section (1000 x 12,500 and 1000 x 25,000,
sparse and dense), and we report timings and memory usage.

See **Timings observed on various systems** section at the end of this
document for `simple_normalize()` and `simple_pca()` times observed
on all our test datasets on various systems.

## PCA on the normalized sparse datasets

```{r}
DelayedArray::setAutoBlockSize(1e8)  # blocks of 100 Mb
```

### 1000 x 12,500 sparse dataset

```{r PCA_sparse1n}
sparse1n <- TENxMatrix(sparse1n_path)
dim(sparse1n)
system.time(pca1s <- simple_PCA(sparse1n))
gc()
```

### 1000 x 25,000 sparse dataset

```{r PCA_sparse2n}
sparse2n <- TENxMatrix(sparse2n_path)
dim(sparse2n)
system.time(pca2s <- simple_PCA(sparse2n))
gc()
```

### About memory usage

With this block size (100 Mb), memory usage (as reported by Unix
command `ps u -p <PID>`, see **Monitoring memory usage** above in
this document) remained < 2.4 Gb at all time.

## PCA on the normalized dense datasets

```{r}
DelayedArray::setAutoBlockSize(1e8)  # blocks of 100 Mb
```

### 1000 x 12,500 dense dataset

```{r PCA_dense1n}
dense1n <- HDF5Array(dense1n_path, name="normalized_counts")
dim(dense1n)
system.time(pca1d <- simple_PCA(dense1n))
gc()
```

### 1000 x 25,000 dense dataset

```{r PCA_dense2n}
dense2n <- HDF5Array(dense2n_path, name="normalized_counts")
dim(dense2n)
system.time(pca2d <- simple_PCA(dense2n))
gc()
```

### About memory usage

With this block size (100 Mb), memory usage (as reported by Unix
command `ps u -p <PID>`, see **Monitoring memory usage** above in
this document) remained < 2.7 Gb at all time.

## Sanity checks

```{r sanity_checks}
stopifnot(all.equal(pca1s, pca1d))
stopifnot(all.equal(pca2s, pca2d))
```


# Timings observed on various systems

Here we report `simple_normalize()` and `simple_pca()` times observed
on all our test datasets on various systems.

## DELL XPS 15 laptop (model 9520)

- **RAM:** 32GB
- **Disk:** 1TB SSD
- **OS:** Linux Ubuntu 24.04
- **Bioconductor/R versions:** 3.21/4.5

### Normalization

<table style="text-align: center;">
  <tr>
    <th></th>
    <th style="background: #CCC; border: 1pt solid #888; padding: 4pt;" colspan="3">
      sparse<br/>(TENxMatrix)
    </th>
    <th style="background: #CCC; border: 1pt solid #888; padding: 4pt;" colspan="3">
      dense<br/>(HDF5Matrix)
    </th>
  </tr>
  <tr style="background: #CCC;">
    <th style="border: 1pt solid #888; padding: 4pt;" rowspan="2">
      object&nbsp;dimensions<br />(genes&nbsp;x&nbsp;cells)
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;" rowspan="2">
      object name
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;" colspan="2">
      block&nbsp;size&nbsp;=&nbsp;250&nbsp;Mb
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;" rowspan="2">
      object name
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;" colspan="2">
      block&nbsp;size&nbsp;=&nbsp;100&nbsp;Mb
    </th>
  </tr>
  <tr style="background: #CCC;">
    <th style="border: 1pt solid #888; padding: 4pt;">
      time (seconds)
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;">
      max. memory<br />used
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;">
      time (seconds)
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;">
      max. memory<br />used
    </th>
  </tr>
  <tr>
    <td style="border: 1pt solid #888; padding: 4pt;">27,998 x 12,500</td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>sparse1</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">40.7</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>dense1</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">52.8</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
  </tr>
  <tr>
    <td style="border: 1pt solid #888; padding: 4pt;">27,998 x 25,000</td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>sparse2</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">85.2</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>dense2</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">111.4</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
  </tr>
  <tr>
    <td style="border: 1pt solid #888; padding: 4pt;">27,998 x 50,000</td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>sparse3</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">172.7</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>dense3</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">223.3</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
  </tr>
  <tr>
    <td style="border: 1pt solid #888; padding: 4pt;">27,998 x 100,000</td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>sparse4</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">346.6</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>dense4</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">456.4</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
  </tr>
  <tr>
    <td style="border: 1pt solid #888; padding: 4pt;">27,998 x 200,000</td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>sparse5</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">742.0</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>dense5</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">942.5</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
  </tr>
</table>

### PCA

<table style="text-align: center;">
  <tr>
    <th></th>
    <th style="background: #CCC; border: 1pt solid #888; padding: 4pt;" colspan="3">
      sparse<br/>(TENxMatrix)
    </th>
    <th style="background: #CCC; border: 1pt solid #888; padding: 4pt;" colspan="3">
      dense<br/>(HDF5Matrix)
    </th>
  </tr>
  <tr style="background: #CCC;">
    <th style="border: 1pt solid #888; padding: 4pt;" rowspan="2">
      object&nbsp;dimensions<br />(genes&nbsp;x&nbsp;cells)
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;" rowspan="2">
      object name
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;" colspan="2">
      block&nbsp;size&nbsp;=&nbsp;100&nbsp;Mb
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;" rowspan="2">
      object name
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;" colspan="2">
      block&nbsp;size&nbsp;=&nbsp;100&nbsp;Mb
    </th>
  </tr>
  <tr style="background: #CCC;">
    <th style="border: 1pt solid #888; padding: 4pt;">
      time (seconds)
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;">
      max. memory<br />used
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;">
      time (seconds)
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;">
      max. memory<br />used
    </th>
  </tr>
  <tr>
    <td style="border: 1pt solid #888; padding: 4pt;">1000 x 12,500</td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>sparse1n</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">47.9</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>dense1n</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">42.2</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
  </tr>
  <tr>
    <td style="border: 1pt solid #888; padding: 4pt;">1000 x 25,000</td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>sparse2n</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">70.0</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>dense2n</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">78.9</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
  </tr>
  <tr>
    <td style="border: 1pt solid #888; padding: 4pt;">1000 x 50,000</td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>sparse3n</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">152.3</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>dense3n</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">176.4</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
  </tr>
  <tr>
    <td style="border: 1pt solid #888; padding: 4pt;">1000 x 100,000</td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>sparse4n</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">289.8</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>dense4n</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">458.7</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
  </tr>
  <tr>
    <td style="border: 1pt solid #888; padding: 4pt;">1000 x 200,000</td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>sparse5n</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">637.4</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>dense5n</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">867.4</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
  </tr>
</table>

## DELL PowerEdge R440 Server

- **RAM:** 128GB
- **Disk:** 1.92TB SSD SATA Mix Use
- **OS:** Linux Ubuntu 24.04
- **Bioconductor/R versions:** 3.21/4.5

### Normalization

Timings coming soon...

### PCA

Timings coming soon...

## Mac Pro (Apple M2 Ultra)

- **RAM:** 192GB
- **Disk:** 2TB SSD
- **OS:** macOS 13.7.1
- **Bioconductor/R versions:** 3.21/4.5

### Normalization

<table style="text-align: center;">
  <tr>
    <th></th>
    <th style="background: #CCC; border: 1pt solid #888; padding: 4pt;" colspan="3">
      sparse<br/>(TENxMatrix)
    </th>
    <th style="background: #CCC; border: 1pt solid #888; padding: 4pt;" colspan="3">
      dense<br/>(HDF5Matrix)
    </th>
  </tr>
  <tr style="background: #CCC;">
    <th style="border: 1pt solid #888; padding: 4pt;" rowspan="2">
      object&nbsp;dimensions<br />(genes&nbsp;x&nbsp;cells)
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;" rowspan="2">
      object name
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;" colspan="2">
      block&nbsp;size&nbsp;=&nbsp;250&nbsp;Mb
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;" rowspan="2">
      object name
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;" colspan="2">
      block&nbsp;size&nbsp;=&nbsp;100&nbsp;Mb
    </th>
  </tr>
  <tr style="background: #CCC;">
    <th style="border: 1pt solid #888; padding: 4pt;">
      time (seconds)
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;">
      max. memory<br />used
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;">
      time (seconds)
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;">
      max. memory<br />used
    </th>
  </tr>
  <tr>
    <td style="border: 1pt solid #888; padding: 4pt;">27,998 x 12,500</td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>sparse1</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">33.6</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>dense1</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">35.3</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
  </tr>
  <tr>
    <td style="border: 1pt solid #888; padding: 4pt;">27,998 x 25,000</td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>sparse2</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">67.4</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>dense2</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">74.2</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
  </tr>
  <tr>
    <td style="border: 1pt solid #888; padding: 4pt;">27,998 x 50,000</td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>sparse3</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">140.3</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>dense3</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">148.1</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
  </tr>
  <tr>
    <td style="border: 1pt solid #888; padding: 4pt;">27,998 x 100,000</td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>sparse4</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">279.9</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>dense4</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">305.4</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
  </tr>
  <tr>
    <td style="border: 1pt solid #888; padding: 4pt;">27,998 x 200,000</td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>sparse5</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">608.1</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>dense5</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">617.8</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
  </tr>
</table>

### PCA

<table style="text-align: center;">
  <tr>
    <th></th>
    <th style="background: #CCC; border: 1pt solid #888; padding: 4pt;" colspan="3">
      sparse<br/>(TENxMatrix)
    </th>
    <th style="background: #CCC; border: 1pt solid #888; padding: 4pt;" colspan="3">
      dense<br/>(HDF5Matrix)
    </th>
  </tr>
  <tr style="background: #CCC;">
    <th style="border: 1pt solid #888; padding: 4pt;" rowspan="2">
      object&nbsp;dimensions<br />(genes&nbsp;x&nbsp;cells)
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;" rowspan="2">
      object name
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;" colspan="2">
      block&nbsp;size&nbsp;=&nbsp;100&nbsp;Mb
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;" rowspan="2">
      object name
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;" colspan="2">
      block&nbsp;size&nbsp;=&nbsp;100&nbsp;Mb
    </th>
  </tr>
  <tr style="background: #CCC;">
    <th style="border: 1pt solid #888; padding: 4pt;">
      time (seconds)
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;">
      max. memory<br />used
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;">
      time (seconds)
    </th>
    <th style="border: 1pt solid #888; padding: 4pt;">
      max. memory<br />used
    </th>
  </tr>
  <tr>
    <td style="border: 1pt solid #888; padding: 4pt;">1000 x 12,500</td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>sparse1n</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">33.7</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>dense1n</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">30.0</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
  </tr>
  <tr>
    <td style="border: 1pt solid #888; padding: 4pt;">1000 x 25,000</td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>sparse2n</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">58.0</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>dense2n</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">56.1</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
  </tr>
  <tr>
    <td style="border: 1pt solid #888; padding: 4pt;">1000 x 50,000</td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>sparse3n</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">117.5</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>dense3n</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">127.9</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
  </tr>
  <tr>
    <td style="border: 1pt solid #888; padding: 4pt;">1000 x 100,000</td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>sparse4n</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">255.1</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>dense4n</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">372.6</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
  </tr>
  <tr>
    <td style="border: 1pt solid #888; padding: 4pt;">1000 x 200,000</td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>sparse5n</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">570.8</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
    <td style="border: 1pt solid #888; padding: 4pt;"><code>dense5n</code></td>
    <td style="border: 1pt solid #888; padding: 4pt;">677.1</td>
    <td style="border: 1pt solid #888; padding: 4pt;"></td>
  </tr>
</table>


# Conclusions

The sparse representation (TENxMatrix) seems to perform slightly better
than the dense representation (HDF5Matrix) when it comes to normalization/PCA
of single cell data. Also the gap in performance between sparse and dense
tends to slightly increase with the size of the dataset.

Normalization and PCA are very roughly linear in time, regardless of
representation (sparse or dense).

Normalization and PCA both perform at almost constant memory, regardless
of representation (sparse or dense).


# Session information

```{r}
sessionInfo()
```

