---
title: "_HDF5Array_ performance"
author:
- name: Hervé Pagès
  affiliation: Fred Hutch Cancer Center, Seattle, WA
date: "Compiled `r doc_date()`; Modified 16 January 2025"
package: HDF5Array
vignette: |
  %\VignetteIndexEntry{HDF5Array performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document
---


# Introduction

The aim of this document is to measure the performance of
the `r Biocpkg("HDF5Array")` package for normalization and PCA
of _on-disk_ single cell data, two computationally intensive
operations at the core of single cell analysis.

The benchmarks presented in this document were specifically designed
to observe the impact of two critical parameters on performance:

1. data representation (i.e. _sparse_ vs _dense_);
2. size of the blocks used for block-processed operations.

Hopefully these benchmarks will also facilitate comparing performance
of single cell analysis workflows based on `r Biocpkg("HDF5Array")`
with workflows based on other tools like Seurat or Scanpy.


# Install and load the required packages

Let's install and load `r Biocpkg("HDF5Array")` as well as the other
packages used in this vignette:
```{r, eval=FALSE}
if (!require("BiocManager", quietly=TRUE))
    install.packages("BiocManager")

pkgs <- c("HDF5Array", "ExperimentHub", "DelayedMatrixStats", "RSpectra")
BiocManager::install(pkgs)
```

Load the packages:
```{r, message=FALSE}
library(HDF5Array)
library(ExperimentHub)
library(DelayedMatrixStats)
library(RSpectra)
```

```{r source_make_timings_table_R, echo=FALSE, results='hide'}
## Needed for the make_timings_table() function.
path <- system.file(package="HDF5Array",
                    "scripts", "make_timings_table.R", mustWork=TRUE)
source(path, verbose=FALSE)
```


# The test datasets

## _Sparse_ vs _dense_ representation

The datasets that we will use for our benchmarks are subsets of the
_1.3 Million Brain Cell Dataset_ from 10x Genomics.

The _1.3 Million Brain Cell Dataset_ is a 27,998 x 1,306,127 matrix
of counts with one gene per row and one cell per column. It's available
via the `r Biocpkg("ExperimentHub")` package in two forms, one that uses
a sparse representation and one that uses a dense representation:
```{r}
hub <- ExperimentHub()
hub["EH1039"]$description  # sparse representation
hub["EH1040"]$description  # dense representation
```

The two datasets are big HDF5 files stored on a remote location.
Let's download them to the local `r Biocpkg("ExperimentHub")` cache
if they are not there yet:
```{r, message=FALSE}
## Note that this will be quick if the HDF5 files are already in the
## local ExperimentHub cache. Otherwise, it will take a while!
full_sparse_h5 <- hub[["EH1039"]]
full_dense_h5  <- hub[["EH1040"]]
```

## TENxMatrix vs HDF5Matrix objects

We use the `TENxMatrix()` and `HDF5Array()` constructors to bring the
sparse and dense datasets in R as DelayedArray derivatives. Note that
this does _not_ load the matrix data in memory.

Bring the sparse dataset in R:
```{r}
## Use 'h5ls(full_sparse_h5)' to find out the group.
full_sparse <- TENxMatrix(full_sparse_h5, group="mm10")
```

Note that `full_sparse` is a 27,998 x 1,306,127 TENxMatrix object:
```{r}
class(full_sparse)
dim(full_sparse)
```
See `?TENxMatrix` in the `r Biocpkg("HDF5Array")` package for more
information about TENxMatrix objects.

Bring the dense dataset in R:
```{r}
## Use 'h5ls(full_dense_h5)' to find out the name of the dataset.
full_dense <- HDF5Array(full_dense_h5, name="counts")
```

Note that `full_dense` is a 27,998 x 1,306,127 HDF5Matrix object that
contains the same data as `full_sparse`:
```{r}
class(full_dense)
dim(full_dense)
```
See `?HDF5Matrix` in the `r Biocpkg("HDF5Array")` package for more
information about HDF5Matrix objects.

Finally note that the dense HDF5 file does not contain the dimnames
of the matrix so we manually add them to `full_sparse`:
```{r}
dimnames(full_dense) <- dimnames(full_sparse)
```

## Create the test datasets

For our benchmarks, below we create subsets of the _1.3 Million Brain
Cell Dataset_ of increasing sizes: subsets with 12,500 cells, 25,000 cells,
50,000 cells, 100,000 cells, and 200,000 cells. Note that subsetting a
TENxMatrix or HDF5Matrix object with `[` is a delayed operation so has
virtually no cost:
```{r}
sparse1 <- full_sparse[ , 1:12500]
dense1  <- full_dense[ , 1:12500]

sparse2 <- full_sparse[ , 1:25000]
dense2  <- full_dense[ , 1:25000]

sparse3 <- full_sparse[ , 1:50000]
dense3  <- full_dense[ , 1:50000]

sparse4 <- full_sparse[ , 1:100000]
dense4  <- full_dense[ , 1:100000]

sparse5 <- full_sparse[ , 1:200000]
dense5  <- full_dense[ , 1:200000]
```


# Block-processed normalization and PCA

## Code used for normalization and PCA

We'll use the following code for normalization:
```{r}
## Keep 1000 most variable genes by default.
simple_normalize <- function(mat, num_variable_genes=1000)
{
    stopifnot(length(dim(mat)) == 2, !is.null(rownames(mat)))
    mat <- mat[rowSums(mat) > 0, ]
    mat <- t(t(mat) * 10000 / colSums(mat))
    row_vars <- rowVars(mat)
    rv_order <- order(row_vars, decreasing=TRUE)
    variable_idx <- head(rv_order, n=num_variable_genes)
    mat <- log1p(mat[variable_idx, ])
    mat / rowSds(mat)
}
```

and the following code for PCA:
```{r}
simple_PCA <- function(mat, k=25)
{
    stopifnot(length(dim(mat)) == 2)
    row_means <- rowMeans(mat)
    Ax <- function(x, args)
        (as.numeric(mat %*% x) - row_means * sum(x))
    Atx <- function(x, args)
        (as.numeric(x %*% mat) - as.vector(row_means %*% x))
    RSpectra::svds(Ax, Atrans=Atx, k=k, dim=dim(mat))
}
```

## Block processing and block size

Note that the implementations of `simple_normalize()` and `simple_PCA()`
are expected to work on any matrix-like object regardless of its exact
type/representation e.g. it can be an ordinary matrix, a SparseMatrix
object from the `r Biocpkg("SparseArray")` package, a dgCMatrix object
from the `r CRANpkg("Matrix")` package, a DelayedMatrix derivative
(TENxMatrix, HDF5Matrix, TileDBMatrix), etc...

However, when the input is a DelayedMatrix object or derivative,
it's important to be aware that:

- Summarization methods like `sum()`, `colSums()`, `rowVars()`, or `rowSds()`,
  and matrix multiplication (`%*%`), are _block-processed_ operations.

- The block size is 100 Mb by default. Increasing or decreasing the block
  size will typically increase or decrease the memory usage
  of _block-processed_ operations. It will also impact performance,
  sometimes in unexpected or counter-intuitive ways.

- The block size can be controlled with `DelayedArray::getAutoBlockSize()`
  and `DelayedArray::setAutoBlockSize()`.

For our benchmarks below, we'll use the following block sizes:

- normalization of the sparse datasets: 250 Mb
- normalization of the dense datasets: 40 Mb
- PCA on the normalized sparse datasets: 40 Mb
- PCA on the normalized dense datasets: 100 Mb (the default)

## Monitoring memory usage

While manually running our benchmarks below on a Linux system, we will
also monitor memory usage at the command line in a terminal with:

    (while true; do ps u -p <PID>; sleep 1; done) >ps.log 2>&1 &

where `<PID>` is the process id of our R session. This will allow us
to measure the maximum amount of memory used by the calls
to `simple_normalize()` or `simple_PCA()`.


# Normalization benchmarks

In this section we run `simple_normalize()` on the two smaller test
datasets only (27,998 x 12,500 and 27,998 x 25,000, sparse and dense),
and we report timings and memory usage.

See **Timings obtained on various systems** section at the end of this
document for `simple_normalize()` and `simple_pca()` timings obtained
on various systems on all our test datasets and using three different
block sizes: 40 Mb, 100 Mb, and 250 Mb.

## Normalizing the sparse datasets

```{r}
DelayedArray::setAutoBlockSize(2.5e8)  # blocks of 250 Mb
```

### 27,998 x 12,500 sparse dataset

```{r normalize_sparse1}
dim(sparse1)
system.time(sparse1n <- simple_normalize(sparse1))
gc()
```

Saving the normalized dataset to a temporary file for PCA later:
```{r, results="hide"}
dim(sparse1n)
sparse1n_path <- tempfile()
writeTENxMatrix(sparse1n, sparse1n_path, group="matrix", level=0)
```

### 27,998 x 25,000 sparse dataset

```{r normalize_sparse2}
dim(sparse2)
system.time(sparse2n <- simple_normalize(sparse2))
gc()
```

Saving the normalized dataset to a temporary file for PCA later:
```{r, results="hide"}
dim(sparse2n)
sparse2n_path <- tempfile()
writeTENxMatrix(sparse2n, sparse2n_path, group="matrix", level=0)
```

### About memory usage

With this block size (250 Mb), memory usage (as reported by Unix
command `ps u -p <PID>`, see **Monitoring memory usage** above in
this document) remained < 3.7 Gb at all time.

## Normalizing the dense datasets

```{r}
DelayedArray::setAutoBlockSize(4e7)  # blocks of 40 Mb
```

### 27,998 x 12,500 dense dataset

```{r normalize_dense1}
dim(dense1)
system.time(dense1n <- simple_normalize(dense1))
gc()
```

Saving the normalized dataset to a temporary file for PCA:
```{r, results="hide"}
dim(dense1n)
dense1n_path <- tempfile()
writeHDF5Array(dense1n, dense1n_path, name="normalized_counts", level=0)
```

### 27,998 x 25,000 dense dataset

```{r normalize_dense2}
dim(dense2)
system.time(dense2n <- simple_normalize(dense2))
gc()
```

Saving the normalized dataset to a temporary file for PCA:
```{r, results="hide"}
dim(dense2n)
dense2n_path <- tempfile()
writeHDF5Array(dense2n, dense2n_path, name="normalized_counts", level=0)
```

### About memory usage

With this block size (40 Mb), memory usage (as reported by Unix
command `ps u -p <PID>`, see **Monitoring memory usage** above in
this document) remained < 2.8 Gb at all time.


# PCA benchmarks

In this section we run `simple_pca()` on the two normalized datasets
obtained in the previous section (1000 x 12,500 and 1000 x 25,000,
sparse and dense), and we report timings and memory usage.

See **Timings obtained on various systems** section at the end of this
document for `simple_normalize()` and `simple_pca()` timings obtained
on various systems on all our test datasets and using three different
block sizes: 40 Mb, 100 Mb, and 250 Mb.

## PCA on the normalized sparse datasets

```{r}
DelayedArray::setAutoBlockSize(4e7)  # blocks of 40 Mb
```

### 1000 x 12,500 sparse dataset

```{r PCA_sparse1n}
sparse1n <- TENxMatrix(sparse1n_path)
dim(sparse1n)
system.time(pca1s <- simple_PCA(sparse1n))
gc()
```

### 1000 x 25,000 sparse dataset

```{r PCA_sparse2n}
sparse2n <- TENxMatrix(sparse2n_path)
dim(sparse2n)
system.time(pca2s <- simple_PCA(sparse2n))
gc()
```

### About memory usage

With this block size (40 Mb), memory usage (as reported by Unix
command `ps u -p <PID>`, see **Monitoring memory usage** above in
this document) remained < 2.4 Gb at all time.

## PCA on the normalized dense datasets

```{r}
DelayedArray::setAutoBlockSize(1e8)  # blocks of 100 Mb
```

### 1000 x 12,500 dense dataset

```{r PCA_dense1n}
dense1n <- HDF5Array(dense1n_path, name="normalized_counts")
dim(dense1n)
system.time(pca1d <- simple_PCA(dense1n))
gc()
```

### 1000 x 25,000 dense dataset

```{r PCA_dense2n}
dense2n <- HDF5Array(dense2n_path, name="normalized_counts")
dim(dense2n)
system.time(pca2d <- simple_PCA(dense2n))
gc()
```

### About memory usage

With this block size (100 Mb), memory usage (as reported by Unix
command `ps u -p <PID>`, see **Monitoring memory usage** above in
this document) remained < 2.7 Gb at all time.

## Sanity checks

```{r sanity_checks}
stopifnot(all.equal(pca1s, pca1d))
stopifnot(all.equal(pca2s, pca2d))
```


# Timings obtained on various systems

Here we report normalization & PCA timings obtained on various systems.
For each system, the results are summarized in a table that shows
the `simple_normalize()` and `simple_pca()` timings obtained on all our test
datasets and using three different block sizes: 40 Mb, 100 Mb, and 250 Mb.
The times with <span style="background: #EFE;">light green</span> background
correspond to the best time amongst the three different block sizes for
a given operation.

## DELL XPS 15 laptop (model 9520)

- **OS:** Linux Ubuntu 24.04
- **Bioconductor/R versions:** 3.21/4.5
- **RAM:** 32GB
- **Disk:** 1TB SSD
- **Disk performance**:
<table><tr><td style="width: 80pt"></td><td>
<pre style="padding: 2pt; background: #EDF; font-size: 8pt;">
    # Output of 'sudo hdparm -Tt &lt;device&gt;':
     Timing cached reads:   35188 MB in  2.00 seconds = 17620.75 MB/sec
     Timing buffered disk reads: 7842 MB in  3.00 seconds = 2613.57 MB/sec
</pre>
</td></tr></table>

```{r xps15_timings, echo=FALSE, results='asis'}
make_timings_table("xps15")
```
_Note: "max. mem. used" columns to be populated soon._

## Supermicro SuperServer 1029GQ-TRT

- **OS:** Linux Ubuntu 22.04
- **Bioconductor/R versions:** 3.21/4.5
- **RAM:** 384GB
- **Disk:** 1.3TB ATA Disk
- **Disk performance**:
<table><tr><td style="width: 80pt"></td><td>
<pre style="padding: 2pt; background: #EDF; font-size: 8pt;">
    # Output of 'sudo hdparm -Tt &lt;device&gt;':
     Timing cached reads:   20592 MB in  1.99 seconds = 10361.41 MB/sec
     Timing buffered disk reads: 1440 MB in  3.00 seconds = 479.66 MB/sec
</pre>
</td></tr></table>

```{r rex3_timings, echo=FALSE, results='asis'}
make_timings_table("rex3")
```
_Note: "max. mem. used" columns to be populated soon._

## Mac Pro (Apple M2 Ultra)

- **OS:** macOS 13.7.1
- **Bioconductor/R versions:** 3.21/4.5
- **RAM:** 192GB
- **Disk:** 2TB SSD
- **Disk performance**: N/A

```{r kjohnson3_timings, echo=FALSE, results='asis'}
make_timings_table("kjohnson3")
```
_Note: "max. mem. used" columns to be populated soon._


# Final remarks

The Supermicro SuperServer 1029GQ-TRT machine is significantly slower
than the other machines. This is most likely due to the less performant
disk.

For PCA, choosing the sparse representation (TENxMatrix) and using small
block sizes (40 Mb) is a clear winner.

For normalization, there's no clear best choice between the parse and dense
representations. More precisely, for this operation, the sparse representation
tends to give the best times when using bigger blocks (250 Mb), whereas the
dense representation tends to give the best times when using smaller blocks
(40 Mb). However, based on the above benchmarks, there's no clear best choice
between "sparse with big blocks" and "dense with small blocks" in terms of
speed. Maybe extending the benchmarks to include more extreme block sizes
(e.g. 20 Mb and 500 Mb) could help break the tie.

Normalization and PCA are roughly linear in time, regardless of
representation (sparse or dense) or block size.

[Needs confirmation] Normalization and PCA both perform at almost
constant memory, regardless of representation (sparse or dense).


# Session information

```{r}
sessionInfo()
```

