---
title: "_HDF5Array_ performance"
author:
- name: Hervé Pagès
  affiliation: Fred Hutch Cancer Center, Seattle, WA
date: "Compiled `r doc_date()`; Modified 14 January 2024"
package: HDF5Array
vignette: |
  %\VignetteIndexEntry{HDF5Array performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document
---


# Introduction

The aim of this short document is to measure the performance of
the `r Biocpkg("HDF5Array")` package for normalization and PCA,
two operations commomly involved in the context of single cell analysis.

The goal is to facilitate comparison with other tools like Seurat
and Scanpy.


# Install and load required packages

Let's install and load `r Biocpkg("HDF5Array")` as well as the other
packages used in this vignette:
```{r, eval=FALSE}
if (!require("BiocManager", quietly=TRUE))
    install.packages("BiocManager")

pkgs <- c("HDF5Array", "ExperimentHub", "DelayedMatrixStats", "RSpectra")
BiocManager::install(pkgs)
```

Load the package:
```{r, message=FALSE}
library(HDF5Array)
library(ExperimentHub)
library(DelayedMatrixStats)
library(RSpectra)
```


# Test datasets

## _Sparse_ vs _dense_ representation

The datasets that we will use in this document are subsets of the
_1.3 Million Brain Cell Dataset_ from 10x Genomics.

The _1.3 Million Brain Cell Dataset_ is a 27,998 x 1,306,127 matrix
of counts with one gene per row and one cell per column. It's available
via the `r Biocpkg("ExperimentHub")` package in two forms, one that uses
a sparse representation and one that uses a dense representation:
```{r}
hub <- ExperimentHub()
hub["EH1039"]$description  # sparse representation
hub["EH1040"]$description  # dense representation
```

The two datasets are big HDF5 files stored on a remote location.
Let's download them to the local `r Biocpkg("ExperimentHub")` cache
if they are not there yet:
```{r, message=FALSE}
## Note that this will be quick if the HDF5 files are already in the
## local ExperimentHub cache. Otherwise, it will take a while
full_sparse_h5 <- hub[["EH1039"]]
full_dense_h5  <- hub[["EH1040"]]
```

## TENxMatrix vs HDF5Matrix object

We use the `TENxMatrix()` and `HDF5Array()` constructors to bring the
sparse and dense datasets in R as DelayedArray derivatives. Note that
this does _not_ load the matrix data in memory.

Bring the sparse dataset in R:
```{r}
## Use 'h5ls(full_sparse_h5)' to find out the group.
full_sparse <- TENxMatrix(full_sparse_h5, group="mm10")
class(full_sparse)
dim(full_sparse)
```

Bring the dense dataset in R:
```{r}
## Use 'h5ls(full_dense_h5)' to find out the name of the dataset.
full_dense <- HDF5Array(full_dense_h5, name="counts")
class(full_dense)
dim(full_dense)
```

Note that the dense HDF5 file does not contain the dimnames of the
matrix so we add them:
```{r}
dimnames(full_dense) <- dimnames(full_sparse)
```

## Create the test datasets

For our benchmarks below we'll use subsets of the _1.3 Million Brain
Cell Dataset_ of increasing sizes: subsets with 10,000 cells, 20,000 cells,
and 40,000 cells:
```{r}
sparse1 <- full_sparse[ , 1:10000]
sparse2 <- full_sparse[ , 1:20000]
sparse3 <- full_sparse[ , 1:40000]

dense1 <- full_dense[ , 1:10000]
dense2 <- full_dense[ , 1:20000]
dense3 <- full_dense[ , 1:40000]
```

# Block-processed normalization and PCA

## Code used for normalization and PCA

We'll use the following code for normalization:
```{r}
## Keep 1000 most variable genes by default.
simple_normalize <- function(mat, num_variable_genes=1000)
{
    stopifnot(length(dim(mat)) == 2, !is.null(rownames(mat)))
    mat <- mat[rowSums(mat) > 0, ]
    mat <- t(t(mat) * 10000 / colSums(mat))
    row_vars <- rowVars(mat)
    rv_order <- order(row_vars, decreasing=TRUE)
    variable_idx <- head(rv_order, n=num_variable_genes)
    mat <- log1p(mat[variable_idx, ])
    mat / rowSds(mat)
}
```

and the following code for PCA:
```{r}
simple_PCA <- function(mat, k=25)
{
    stopifnot(length(dim(mat)) == 2)
    row_means <- rowMeans(mat)
    Ax <- function(x, args)
        (as.numeric(mat %*% x) - row_means * sum(x))
    Atx <- function(x, args)
        (as.numeric(x %*% mat) - as.vector(row_means %*% x))
    RSpectra::svds(Ax, Atrans=Atx, k=k, dim=dim(mat))
}
```

## Block processing and block size

Note that the implementations of `simple_normalize()` and `simple_PCA()`
are expected to work on any matrix-like object regardless of its exact
type/representation e.g. it can be an ordinary matrix, a SparseMatrix
object from the `r Biocpkg("SparseArray")` package, an dgCMatrix object
from the `r Biocpkg("Matrix")` package, a DelayedMatrix derivative
(TENxMatrix, HDF5Matrix, TileDBMatrix), etc...

However, when the input is a DelayedMatrix object or derivative,
it's important to be aware that:

- Summarization methods like `sum()`, `colSums()`, `rowVars()`, or `rowSds()`,
  and matrix multiplication (`%*%`), are _block-processed_ operations.

- The block size is 100 Mb by default. Increasing or decreasing the block
  size will increase or decrease the memory usage of _block-processed_
  operations.

- The block size can be controlled with `DelayedArray::getAutoBlockSize()`
  and `DelayedArray::setAutoBlockSize()`.

For out benchmarks below, we'll use the following block sizes:
- normalization of the sparse datasets: 250 Mb
- normalization of the dense datasets: 100 Mb
- PCA on the normalized sparse datasets: 100 Mb
- PCA on the normalized dense datasets: 100 Mb

## Monitoring memory usage

While manually running our benchmarks below on a Linux system, we monitored
memory usage at the command line in a terminal with:

    (while true; do ps u -q <PID>; sleep 1; done) >ps.log 2>&1 &

where `<PID>` was the process id of our R session. This allowed us to measure
the maximum amount of memory used by the calls to `simple_normalize()`
or `simple_PCA()`.


# Normalization benchmarks

We report timings and memory usage.

## Normalizing the sparse datasets

```{r}
DelayedArray::setAutoBlockSize(2.5e8)  # blocks of 250 Mb
```

### 27,998 x 10,000 sparse dataset

```{r normalize_sparse1}
dim(sparse1)
system.time(sparse1n <- simple_normalize(sparse1))
gc()
```

Saving the normalized dataset to a temporary file for PCA later:
```{r, results="hide"}
dim(sparse1n)
sparse1n_path <- tempfile()
writeTENxMatrix(sparse1n, sparse1n_path, group="matrix", level=0)
```

### 27,998 x 20,000 sparse dataset

```{r normalize_sparse2}
dim(sparse2)
system.time(sparse2n <- simple_normalize(sparse2))
gc()
```

Saving the normalized dataset to a temporary file for PCA later:
```{r, results="hide"}
dim(sparse2n)
sparse2n_path <- tempfile()
writeTENxMatrix(sparse2n, sparse2n_path, group="matrix", level=0)
```

### 27,998 x 40,000 sparse dataset

```{r normalize_sparse3}
dim(sparse3)
system.time(sparse3n <- simple_normalize(sparse3))
gc()
```

Saving the normalized dataset to a temporary file for PCA later:
```{r, results="hide"}
dim(sparse3n)
sparse3n_path <- tempfile()
writeTENxMatrix(sparse3n, sparse3n_path, group="matrix", level=0)
```

### About memory usage

With this block size (250 Mb), memory usage (as reported by Unix
command `ps u -q <PID>`, see **Monitoring memory usage** above in
this document) remained < 3.7 Gb at all time.

## Normalizing the dense datasets

```{r}
DelayedArray::setAutoBlockSize(1e8)  # blocks of 100 Mb
```

### 27,998 x 10,000 dense dataset

```{r normalize_dense1}
dim(dense1)
system.time(dense1n <- simple_normalize(dense1))
gc()
```

Saving the normalized dataset to a temporary file for PCA:
```{r, results="hide"}
dim(dense1n)
dense1n_path <- tempfile()
writeHDF5Array(dense1n, dense1n_path, name="normalized_counts", level=0)
```

### 27,998 x 20,000 dense dataset

```{r normalize_dense2}
dim(dense2)
system.time(dense2n <- simple_normalize(dense2))
gc()
```

Saving the normalized dataset to a temporary file for PCA:
```{r, results="hide"}
dim(dense2n)
dense2n_path <- tempfile()
writeHDF5Array(dense2n, dense2n_path, name="normalized_counts", level=0)
```

### 27,998 x 40,000 dense dataset

```{r normalize_dense3}
dim(dense3)
system.time(dense3n <- simple_normalize(dense3))
gc()
```

Saving the normalized dataset to a temporary file for PCA:
```{r, results="hide"}
dim(dense3n)
dense3n_path <- tempfile()
writeHDF5Array(dense3n, dense3n_path, name="normalized_counts", level=0)
```

### About memory usage

With this block size (100 Mb), memory usage (as reported by Unix
command `ps u -q <PID>`, see **Monitoring memory usage** above in
this document) remained < 2.8 Gb at all time.


# PCA benchmarks

We report timings and memory usage.

## PCA on the normalized sparse datasets

```{r}
DelayedArray::setAutoBlockSize(1e8)  # blocks of 100 Mb
```

### 1000 x 10,000 sparse dataset

```{r PCA_sparse1n}
sparse1n <- TENxMatrix(sparse1n_path)
dim(sparse1n)
system.time(pca1s <- simple_PCA(sparse1n))
gc()
```

### 1000 x 20,000 sparse dataset

```{r PCA_sparse2n}
sparse2n <- TENxMatrix(sparse2n_path)
dim(sparse2n)
system.time(pca2s <- simple_PCA(sparse2n))
gc()
```

### 1000 x 40,000 sparse dataset

```{r PCA_sparse3n}
sparse3n <- TENxMatrix(sparse3n_path)
dim(sparse3n)
system.time(pca3s <- simple_PCA(sparse3n))
gc()
```

### About memory usage

With this block size (100 Mb), memory usage (as reported by Unix
command `ps u -q <PID>`, see **Monitoring memory usage** above in
this document) remained < 2.4 Gb at all time.

## PCA on the normalized dense datasets

```{r}
DelayedArray::setAutoBlockSize(1e8)  # blocks of 100 Mb
```

### 1000 x 10,000 dense dataset

```{r PCA_dense1n}
dense1n <- HDF5Array(dense1n_path, name="normalized_counts")
dim(dense1n)
system.time(pca1d <- simple_PCA(dense1n))
gc()
```

### 1000 x 20,000 dense dataset

```{r PCA_dense2n}
dense2n <- HDF5Array(dense2n_path, name="normalized_counts")
dim(dense2n)
system.time(pca2d <- simple_PCA(dense2n))
gc()
```

### 1000 x 40,000 dense dataset

```{r PCA_dense3n}
dense3n <- HDF5Array(dense3n_path, name="normalized_counts")
dim(dense3n)
system.time(pca3d <- simple_PCA(dense3n))
gc()
```

### About memory usage

With this block size (100 Mb), memory usage (as reported by Unix
command `ps u -q <PID>`, see **Monitoring memory usage** above in
this document) remained < 2.7 Gb at all time.

## Sanity checks

```{r sanity_checks}
stopifnot(all.equal(pca1s, pca1d))
stopifnot(all.equal(pca2s, pca2d))
stopifnot(all.equal(pca3s, pca3d))
```


# Conclusions

There's no significant difference in performance between sparse (TENxMatrix)
and dense (HDF5Matrix) representations when it comes to normalization/PCA.

Normalization and PCA are linear in time, regardless of representation
(sparse or dense).

Both operations perform at constant memory.

# Session information

```{r}
sessionInfo()
```

